---
title: "Data Analysis Project - Group C63"
author: "Daniel Shevchuk, Katie Giamarese, Josh Kim, Vithuran Rathakrishnan"
date: 'November 22, 2024'
output:
  html_document: 
    theme: readable
    toc: yes
  pdf_document: default
urlcolor: cyan
---

***

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
options(scipen = 1, digits = 4, width = 80, fig.align = "center")
```

## Introduction

The dataset we are going to be analyzing in this project is from the Sloan Digitial Sky Survey (SDSS). The SDSS is long term multi-spectral imaging and spectroscopic redshift survey using a dedicated 2.5 meter optical telescope at Apache Point Observatory in New Mexico. The project began in 2000 and has run four separate generations of observations with a fifth currently underway. Data is released periodically throughout each phase with the most recent data release 18, of which a subset of 100,000 observations we are using as our dataset, being made public in January 2023.

The primary reason we selected this dataset is for interest in the scientific research being conducted using this data. The SDSS website lists the data being used in subjects covering topics such as distant quasars at the limits of the observable universe, the distribution of galaxies within the observable universe, and creating three dimensional maps of stars in our galaxy.

For the purposes of this report, we will be focusing specifically on the variables: u, g, r, i, z, petroRad_u, petroRad_g, petroRad_r, petroRad_i, petroRad_z and the categorical class with values "STAR", "GALAXY", and "QUASAR". We will be using these variables to predict the redshift of any particular object. 

### Variable descriptions

The redshift of an object is how "shifted" the photon is to longer wavelengths and is measured as a fractional value between 0 and 1 of how much the wavelength has been increased. The variables u, g, r, i, z correspond to the raw brightness of each object in the five photometric bands; namely ultraviolet band, green band, red band, infrared band, and near infrared band respectively. 

We must note that units for the u, g, r, i and z variables are the apparent brightness of a celestial object and are measured in apparent magnitude, formally called magnitudes per square arcsecond (MPSAS). An arcsecond is 1/3600 of a degree (the entire sky is considered to be 360 degrees) and the way the measurement is calculated dictates that the brighter the object, the lower the number assigned as a magnitude.

The petroRad variables correspond to the petrosian radius, which measures how the brightness of an object varies with distance from its center. The value is defined as the distance from the object's center where the ratio of the local surface brightness to the average surface brightness reaches a certain predefined value (set to a magnitude of 0.2 in the case of this dataset). In essence, using the petrosian radius in this way "normalizes" the size of an object based on its average brightness, since objects further away appear dimmer and smaller than closer objects.

## Methods

We first read in our data, and select the appropriate columns which we are going to be using for our models:

```{r, warning = FALSE}
data = read.csv("SDSS_DR18.csv")
df = data[c("u", "g", "r", "i", "z", "petroRad_u", "petroRad_g", "petroRad_r", "petroRad_i", "petroRad_z", "class", "redshift")]
head(df)
```
Now we fit our first simple linear model, which just uses all the variables as predictors with no further modification:
```{r}
initial_model <- lm(redshift ~ ., data = df)
cat("Initial Model Summary:\n")
summary(initial_model)
```

Now that we have a baseline model we try a much larger model, specifically using all of the base predictors as well as their second order interactions terms with each other:

```{r}
interaction_model = lm(redshift ~ .^2, data = df)

cat("\nInteraction Model Summary:\n")
summary(interaction_model)
```
We find that the $R^2$ value has improved, however, a massive downside of this improvement in the model is the massive increase in number of predictors for only a marginal improvement in explanatory power of about 4%. Where the previous simple model had ten numeric predictors with two factor levels of a categorical variable, this interaction model has 78 including the intercept term. 

This model is clearly too complicated to be of practical use, but there are several terms which have very large p-values indicating that they do not add much to the predictive power of the interaction model. We can instead use the method of forwards/backwards search by use of AIC and BIC criteria in order to reduce the model complexity hopefully without sacrificing too much accuracy.


# TO DO: perform fwd/back selection using null model as start instead of full interaction model
# FYI: This code block takes 10 minutes to run
# Also need to note stepwise is not very good for selecting a model, LASSO or Elastic Net would be a better algorithm for removing predictors

null_model <- lm(redshift ~ 1, data = df)
full_model <- lm(redshift ~ ., data = df)

```{r}
library(MASS)

# Selection using AIC criteria
final_model_aic <- step(interaction_model, direction = "both", trace = FALSE)
cat("\nFinal Model (After Stepwise Selection) Summary:\n")
summary(final_model_aic)

# Selection using BIC criteria
logn = log(length(resid(interaction_model)))
model_bic = step(interaction_model, direction = "both", k = logn, trace = FALSE)
summary(model_bic)

null_model = lm(redshift ~ 1, data = df)
null_model_aic = step(null_model, direction = "both", scope = list(upper = full_model), trace = FALSE)
summary(null_model_aic)


cat("\nAIC and BIC Comparison:\n")

```{r}
# Compare AIC of each model
cat("\nAIC for Initial Model: ", AIC(initial_model), "\n")
cat("BIC for Initial Model: ", BIC(initial_model), "\n")
cat("R Squared for Initial Model: ", summary(initial_model)$r.squared, "\n")

cat("\nAIC for Interaction Model: ", AIC(interaction_model), "\n")
cat("BIC for interaction model: ", BIC(interaction_model), "\n")
cat("R Squared for Interaction Model: ", summary(interaction_model)$r.squared, "\n")

cat("\nAIC for AIC selected model: ", AIC(final_model_aic), "\n")
cat("BIC for AIC selected model: ", BIC(final_model_aic), "\n")
cat("R Squared for AIC Stepwise Model: ", summary(final_model_aic)$r.squared, "\n")

cat("\nAIC for BIC selected model: ", AIC(model_bic), "\n")
cat("BIC for BIC selected model: ", BIC(model_bic), "\n")
cat("R Squared for AIC Stepwise Model: ", summary(model_bic)$r.squared)
```

```{r}
# Determine the best model based on the lowest AIC
aic_values <- c(AIC(initial_model), AIC(interaction_model), AIC(final_model_aic))
model_names <- c("Initial Model", "Interaction Model", "Final Model (Stepwise)")
best_model_index <- which.min(aic_values)
cat("\nThe best model is: ", model_names[best_model_index])
```


## Results

                Model         AIC         BIC
1              Initial   -438.3181   -305.1372
2          Interaction -15619.1862 -14867.6650
3    Forward Selection   -444.6765   -349.5473
4 Backward Elimination   -444.6765   -349.5473
 



AIC measures model quality, balancing fit and model complexity. Lower AIC values indicate a better model because it achieves a better fit with fewer predictors or unnecessary complexity. Interaction Model has the lowest AIC and BIC, the complexity introduced by the interactions might make it less desirable if generalization is a priority. The forward/backward elimination models strike a balance between complexity and fit, and they have slightly better AIC and BIC than the initial model, without the additional complexity of interaction terms


## Discussion

The final model captures the relationships between the photometric properties (u, g, r, i, z) and petrosian radii (petroRad_u, petroRad_g, petroRad_r, petroRad_i, petroRad_z) 
of celestial objects and their redshift. The redshift of a celestial object is a critical measurement in astrophysics, reflecting the degree to which the object's light has shifted toward longer wavelengths. Our model provides a way to predict redshift using data and available features from the Sloan Digital Sky Survey (SDSS). The final model provides an efficient method for predicting redshift based on photometric and structural data. This can be valuable in astrophysical studies because it allows researchers to quickly estimate redshift without relying on time-intensive spectroscopic observations.


## Appendix
