---
title: "Data Analysis Project - Group C63"
author: "Daniel Shevchuk, Katie Giamarese, Josh Kim, Vithuran Rathakrishnan"
date: 'November 22, 2024'
output:
  html_document: default
    theme: readable
    toc: yes
  pdf_document:
urlcolor: cyan
---

***

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
options(scipen = 1, digits = 4, width = 80, fig.align = "center")
```

## Introduction

The dataset we are going to be analyzing in this project is from the Sloan Digitial Sky Survey (SDSS). The SDSS is long term multi-spectral imaging and spectroscopic redshift survey using a dedicated 2.5 meter optical telescope at Apache Point Observatory in New Mexico. The project began in 2000 and has run four separate generations of observations with a fifth currently underway. Data is released periodically throughout each phase with the most recent data release 18, of which a subset of 100,000 observations we are using as our dataset, being made public in January 2023.

The primary reason we selected this dataset is for interest in the scientific research being conducted using this data. The SDSS website lists the data being used in subjects covering topics such as distant quasars at the limits of the observable universe, the distribution of galaxies within the observable universe, and creating three dimensional maps of stars in our galaxy.

For the purposes of this report, we will be focusing specifically on the variables: u, g, r, i, z, petroRad_u, petroRad_g, petroRad_r, petroRad_i, petroRad_z and the categorical class with values "STAR", "GALAXY", and "QUASAR". We will be using these variables to predict the redshift of any particular object. 

### Variable descriptions

The redshift of an object is how "shifted" the photon is to longer wavelengths and is measured as a fractional value between 0 and 1 of how much the wavelength has been increased. The variables u, g, r, i, z correspond to the raw brightness of each object in the five photometric bands; namely ultraviolet band, green band, red band, infrared band, and near infrared band respectively. 

We must note that units for the u, g, r, i and z variables are the apparent brightness of a celestial object and are measured in apparent magnitude, formally called magnitudes per square arcsecond (MPSAS). An arcsecond is 1/3600 of a degree (the entire sky is considered to be 360 degrees) and the way the measurement is calculated dictates that the brighter the object, the lower the number assigned as a magnitude.

The petroRad variables correspond to the petrosian radius, which measures how the brightness of an object varies with distance from its center. The value is defined as the distance from the object's center where the ratio of the local surface brightness to the average surface brightness reaches a certain predefined value (set to a magnitude of 0.2 in the case of this dataset). In essence, the petrosian radius "normalizes" the size of an object based on its average brightness, since objects further away appear dimmer and smaller than closer objects.

## Methods

```{r, warning = FALSE}
data = read.csv("SDSS_DR18.csv")
df = data[["u", "g", "r", "i", "z", "petroRad_u", "petroRad_g", "petroRad_r", "petroRad_i", "petroRad_z", "redshift"]]
head(df)
```

# I literally have no idea if this is what we were supposed to do these instructions are so unclear
# I tried absolutely feel free to change anything and everything


# ----------------------------
# Model 1: Initial Linear Model
# ----------------------------
initial_model <- lm(redshift ~ ., data = df)
cat("Initial Model Summary:\n")
summary(initial_model)

# ----------------------------
# Model 2: Adding Interaction Terms
# ----------------------------
interaction_model <- lm(
  redshift ~ (u + g + r + i + z) * (petroRad_u + petroRad_g + petroRad_r + petroRad_i + petroRad_z),
  data = df
)
cat("\nInteraction Model Summary:\n")
summary(interaction_model)

# ----------------------------
# Model 3: Model Selection
# ----------------------------
library(MASS)
final_model <- stepAIC(interaction_model, direction = "both")
cat("\nFinal Model (After Stepwise Selection) Summary:\n")
summary(final_model)


# Compare AIC of each model
cat("\nAIC for Initial Model: ", AIC(initial_model), "\n")
cat("AIC for Interaction Model: ", AIC(interaction_model), "\n")
cat("AIC for Final Model (After Stepwise Selection): ", AIC(final_model), "\n")

# Determine the best model based on the lowest AIC
aic_values <- c(AIC(initial_model), AIC(interaction_model), AIC(final_model))
model_names <- c("Initial Model", "Interaction Model", "Final Model (Stepwise)")
best_model_index <- which.min(aic_values)
cat("\nThe best model is: ", model_names[best_model_index])


### Results

AIC for Initial Model:  81186.36 
AIC for Interaction Model:  75952.31 
AIC for Final Model (After Stepwise Selection):  75949.73 
The best model is:  Final Model (Stepwise)

AIC measures model quality, balancing fit and model complexity.
Lower AIC values indicate a better model, because it achieves a better fit with fewer predictors or unnecessary complexity.
The Final Model has the lowest AIC of the three models, indicating that it fits the data well
It excludes predictors or interactions that do not contribute significantly to the model.

## Discussion

## Appendix
